The Opioid Crisis: State Regulations and Labor Market Outcomes \\ [2ex] \large CEMFI

### 2024-06-17 slides/paper.tex
replication

% Pablo Muñoz: clarificar a que me refiero con binding

@@

For the past two decades, the United States has been facing the worst public health crisis in its recent history.
The so-called opioid epidemic, characterized by a rising number of deaths due to opioid overdose, has caused more than 680,000 fatalities between 1999 and 2022 (\cite{CDCMCD9920}, \cite{CDCMCD1822}), mostly among non-college-educated Americans.
The sheer magnitude of this crisis has left a significant footprint on the labor market, prompting economists to study the relationship between opioid misuse and labor market outcomes.
However, little is known about how labor market regulations have shaped this relationship.
Minimum wage rates emerge as a natural candidate for studying this dynamic since they affect the same subpopulation of low-educated workers and have increased with varying intensity during the same period.

@@

This paper argues that minimum wage rates affect the relationship between opioid misuse and labor market outcomes.
To measure this relationship, I use the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states.
These programs require authorized users to check a patient's prescription history before prescribing opioids.
By effectively combating opioid misuse, these measures force those addicted to either quit opioid consumption or switch to illegal alternatives.
The incentives to choose one option over the other depend on the labor market opportunities available to those who withdraw. 
Quitting opioids, although costly and often painful, offers benefits not only in better health outcomes but also in job prospects. 
However, for those facing a minimum wage high enough to lower labor demand (binding), the benefits of quitting opioids might be lower than for those with greater job opportunities.
My estimations support this intuition, showing increases in labor force participation and unemployment rates as a result of prescription policies only where the minimum wage was not binding, while a surge in heroin consumption occurred where it was.

@@

% Condensate the whole background part into a single (two) paragraph
These findings contribute to the growing body of literature studying the economic effects of the opioid crisis.
Among this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation.
This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
While well-intentioned, many of these policies have had unintended consequences, such as the rise in heroin consumption following the reformulation of OxyContin.
Prescription Drug Monitoring Programs (PDMPs), however, have been less scrutinized, with recent research indicating that, in their more stringent form, they can effectively reduce prescription opioid misuse.

Yet, how heterogenous these effects are across labor markets remains an open question.
Different incentives can lead victims of opioid misuse to take different decisions regarding both their health and their economic prospects. 
Labor market regulations, by altering the labor outcomes of workers, influence this decision-making process by affecting the incentives to quit opioids -- a family of drugs that can severely impair the ability to work.
To understand how these regulations have shaped the ongoing crisis, I focus on minimum wage rates, which have increased in all states at different speeds during the same period as the opioid crisis, thereby affecting the labor market opportunities of low-educated workers.
Critical to our argument is not the rates themselves, but how binding they are across different labor markets: the more binding the minimum wage, the less incentives workers have to quit opioids, as they face a tougher job market search.

I use several data sources to obtain a distribution of coefficients of the impact of opioid misuse on labor market outcomes across different minimum wage bindingness levels.
The US Bureau of Labor Statistics (BLS) LAUS database provides county-level labor market statistics, while the US Census Bureau's Population Estimates Program (PEP) offers county-level population and demographic data.
From \textcite{Horwitz2021} I obtain data on Prescription Drug Monitoring Programs (PDMPs). 
The BLS's Occupational Employment and Wage Statistics (OEWS) database provides national industry-specific occupations and wages statistics, \textcite{Eckert2020} offers county-level industry-specific employment data, and the US Department of Labor's Changes in Basic Minimum Wages provides information on minimum wage rates; all of these data are relevant for constructing a measure of minimum wage bindingness.
The Centers for Disease Control (CDC) National Vital Statistics System offers data on total drug poisoning mortality by county, while the CDC's Multiple Cause of Death database provides overdose deaths at the state-year level for different drugs.
Finally, the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) offers opioid distribution data.
The joint panel covers the United states for the period 2003-2016.

@@

\textbf{Contribution}: shedding light on how the labor effects of the Opioid Crisis have been affected by minimum wage rates

Interest in how labor market regulations interact with Substance Abuse Disorders

These results shed light on how labor market regulations in general, and minimum wage rates in particular, give rise to heterogeneities in the effects of the opioid crisis on the labor market.
Moreover, they highlight the importance of blah blah.

OUTLINE of the paper (the paper is structured as follows...)

@@

In this section, I provide a summary of how the opioid crisis started and what effects it had on the labor market in general. 
Then, I discuss some of the policies implemented to curb the opioid misuse, and how some of them resulted ineffective or simply backfired. 
Finally, I describe how labor market incentives, and in particular minimum wage rates, can explain some of these policy failures and the heterogeneity of the labor market outcomes.

@@

% CHECKED
Although the significance of the impact of minimum wages on labor demand has been questioned since \textcite{Card1994} published their seminal paper, recent research shows that minimum wage increases can have negative effects on employment in certain contexts.
For instance, \textcite{Cengiz2019} found that minimum wage increases in the United States led to a decrease in employment in tradeable sectors.
This finding is consistent with \textcite{Harasztosi2019}, who observed that the impact on employment is larger in industries where passing costs to consumers is more difficult, using data from Hungary.
\textcite{Clemens2019} showed that minimum wage increases in the United States led to a decrease in employment in states bound by the federal minimum wage increase during the Great Recession.
Furthermore, \textcite{Clemens2021} showed that large increases in minimum wages have reduced employment rates among individuals with low levels of education, in accordance with \textcite{Neumark2004},although the effect is negligible for small increases.
In a recent paper, \textcite{Engbom2022} pointed out that the effect of these increases on employment can be muted by reallocations of workers across firms.
Crucially, they argue that lower-productivity firms cut vacancy creation and hiring, which in our context could lead to lower job finding rates where the minimum wage is more binding.

@@

From (1), we see that the determinants of the decision to quit are key to understand what happens when a prescription law is passed.
This will, in turn, depend on the labor market opportunities the individual faces.
The addicted individual faces a labor market with a job finding rate

@@

The decision to quit opioids is hence affected by the minimum wage in the following way:
If the individual quits, she will look for a job which returns an utility \(u(w_{it}L_{it})\), where \(w_{it}\) is the wage she receives and \(L_{it}\) is the number of hours she works.
The probability of finding a job is given by the job finding rate.
If the product of the two is greater than the disutility of quitting, \(\mathcal{C}^{quit}\), the individual will quit.
Formally

@@

Unless a corner solution exists (i.e, the disutility of quitting is greater or lower than the maximum utility the individual can get from working), there must be some wage level such that

@@

that is, for some wage level, the job finding rate is such that the utility from working is greater than the disutility of quitting.
Notice that this is irrespective of the differences in levels of wages, since the job finding rate is also playing a role.

If condition (6) holds, after a prescription law is passed (\(z_{t} = 1\)) we should see an increase in participation rates where the minimum wage is less binding (\(O_{it} = 0\)) but also an increase in illegal opioid consumption where the minimum wage is more binding (\(O_{it} = h_{it}\)).
We test these hypotheses in the next sections.

@@

To uncover the heterogeneity in the effects of the opioid crisis on the labor market created by minimum wage rates, I use several data sources.
The result is a panel of county-level data for the United States for the period 2003-2016 that includes labor, demographic and health variables.
Where monthly data is not available, I use the data available for the year.
The same holds for the minimum wage rates, which are set at the state level but are inputed to the counties.

\textit{Labor market data and demographics}: I use the US Bureau of Labor Statistics (BLS) Local Area Unemployment Statistics (LAUS) database to obtain county-level labor market statistics. 
In particular, I obtain monthly data on labor force participation, employment, and unemployment rates at the county level.
The dataset covers the period 1990-2023. US Census Bureau's Population Estimates Program (PEP) I am able to retrieve demograpic shares at the county-year level, from 1969 to 2020.
The data gives me the population of each area for each gender, race and origin (hispainc or non-hispanic).
Moreover, I can obtain the number of people in each of the 19 age groups considered, which is used to construct the labor force participation rate.

\textit{Minimum wage bindingness}: I use the US Department of Labor's Changes in Basic Minimum Wages to obtain information on minimum wage rates at the state-level, from 1968 to 2023.
Estimates of the wage distribution at the county level are requiered to construct a measure of minimum wage bindingness, which is outlined in the next section.
For this purpose, I use the US Bureau of Labor Statistics (BLS) Occupational Employment and Wage Statistics (OEWS) database to obtain national industry-specific occupations and wages statistics.
This dataset reports the number of workers in each occupation and the 10th, 25th, median, 75th and 90th wage percentiles for each industry, from 1997 to 2023.
Moreover, I use the data from \textcite{Eckert2020} to obtain county-level industry-specific employment data, from 1975 to 2016.
These last two datasets have a yearly frequency. 4-digit NAICS codes are used to match the industries across datasets.

\textit{Prescription Drug Monitoring Programs (PDMPs)}: I use data from \textcite{Horwitz2021} to obtain information on Prescription Drug Monitoring Programs (PDMPs) from 1990 to 2019.
The data is at the state-year level and reports the type of PDMP implemented in each state.
Table 1 shows the timing of the implementation of Modern System (the last two columns) and Must Query (the first two) PDMPs across states.
A Modern System PDMP is one where the electronic database becomes accessible to any authorized user, while a Must Query PDMP is one where the law mandates the prescriber to check the database before prescribing a listed opioid.
This last type of PDMP is the one that is expected to have the most impact on opioid misuse, and the one I employ in the estimations.

\textit{Prescriptions and overdose deaths}: I use data from the Centers for Disease Control (CDC) National Center for Health Statistics (NCHS) to obtain information on total drug poisoning mortality by county, from 2003 to 2021.
To avoid suppression of data, \textcite{CLMort} employs Hierarchical Bayesian models with spatial and temporal random effects to estimate the number of deaths where the data is not available.
I further use the Multiple Cause of Death database (\cite{CDCMCD9920}), which reports overdose deaths at the state-year level for different drugs, from 1999 to 2020, to obtain information on overdose deaths related to heroin (T.40.1), methadone (T.40.2), other opioids (T.40.3) and other synthetic narcotics (T.40.4).
The codes reffer to the Multiple Causes of Death ICD-10 classification. 
Overdose deaths are identified using the Underlying Cause of Death ICD-10 codes X40-X44, X60-X64, X85, and Y10-Y14.

Finally, I use the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) to obtain opioid distribution data, from 2006 to 2014.
This dataset records the distribution of all Schedule II substances by active ingredient at the 3-digit ZIP code level.
The data is aggregated at the county-month level.
The shorter time span does not affect the main specifications, since I use prescription rates only to assure that the PDMPs are effective.
% Schedule II -> oxycodone and hydrocodone

@@

@@

@@

\section*{Conclusions}

@@

>>>>

@@

For the past two decades, the United States has been facing the worst public health crisis in its recent history.
The so-called opioid epidemic, characterized by a rising number of deaths due to opioid overdose, has caused more than 680,000 fatalities between 1999 and 2022 (\cite{CDCMCD9920}, \cite{CDCMCD1822}), mostly among non-college-educated Americans.
The sheer magnitude of this crisis has significantly impacted the labor market, prompting economists to study the relationship between opioid misuse and labor market outcomes.
However, little is known about how labor market regulations have shaped this relationship.
Minimum wage rates emerge as a natural candidate for studying this dynamic since they affect the same subpopulation of low-educated workers and have increased with varying intensity during the same period.

@@

\begin{comment}
    This paper argues that minimum wage rates affect the relationship between opioid misuse and labor market outcomes.
    To measure this relationship, I use the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states.
    These programs require authorized users to check a patient's prescription history before prescribing opioids.
    By effectively combating opioid misuse, these measures force those addicted to either quit opioid consumption or switch to illegal alternatives.
    The incentives to choose one option over the other depend on the labor market opportunities available to those who withdraw. 
    Quitting opioids, although costly and often painful, offers benefits not only in better health outcomes but also in job prospects. 
    However, for those facing a minimum wage high enough to lower labor demand (binding), the benefits of quitting opioids might be lower than for those with greater job opportunities.
    My estimations support this intuition, showing increases in labor force participation and unemployment rates as a result of prescription policies only where the minimum wage was not binding, while a surge in heroin consumption occurred where it was.
\end{comment}

This paper argues that minimum wage rates influence the relationship between opioid misuse and labor market outcomes.
I measure this relationship using the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, which require authorized users to check a patient's prescription history before prescribing opioids. 
These programs force addicts to either quit opioids or switch to illegal alternatives, with the choice depending on available labor market opportunities.
Quitting opioids, though difficult, improves health and job prospects.
However, in areas with high (binding) minimum wages, reduced labor demand might lower the benefits of quitting.
My estimations show increases in labor force participation and unemployment rates only where the minimum wage was not binding, while a surge in heroin consumption occurred where it was.

@@

% Condensate the whole background part into a single (two) paragraph maybe not mention lit yet

These findings contribute to the growing body of literature studying the economic effects of the opioid crisis.
Within this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation.
This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
While well-intentioned, many of these policies have had unintended consequences, such as the rise in heroin consumption following the reformulation of OxyContin.
However, Prescription Drug Monitoring Programs (PDMPs), have been less scrutinized.
Recent research indicates that, in their more stringent form, these programs can effectively reduce prescription opioid misuse.

\begin{comment}
    These findings contribute to the growing body of literature studying the economic effects of the opioid crisis and of the policies implemented to curb it.
    Within this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation.
    This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
    Prescription Drug Monitoring Programs (PDMPs) 
    Recent research indicates that, in their more stringent form, these programs can effectively reduce prescription opioid misuse.
\end{comment}

Yet, how heterogenous these effects are across labor markets remains an open question.
Different incentives can lead victims of opioid misuse to take different decisions regarding both their health and their economic prospects. 
Labor market regulations, by altering the labor outcomes of workers, influence this decision-making process by affecting the incentives to quit opioids -- a family of drugs that can severely impair the ability to work.
To understand how these regulations have shaped the ongoing crisis, I focus on minimum wage rates, which have increased in all states at different speeds during the same period as the opioid crisis, thereby affecting the labor market opportunities of low-educated workers.
Critical to our argument are not the rates themselves, but how binding they are across different labor markets: the more binding the minimum wage, the less incentives workers have to quit opioids, as they face a tougher job market search.

I use several data sources to obtain a distribution of coefficients of the impact of opioid misuse on labor market outcomes across different minimum wage bindingness levels.
The US Bureau of Labor Statistics (BLS) LAUS database provides county-level labor market statistics, while the US Census Bureau's Population Estimates Program (PEP) offers county-level population and demographic data.
From \textcite{Horwitz2021} I obtain data on Prescription Drug Monitoring Programs (PDMPs). 
The BLS's Occupational Employment and Wage Statistics (OEWS) database provides national industry-specific occupations and wages statistics, \textcite{Eckert2020} offers county-level industry-specific employment data, and the US Department of Labor's Changes in Basic Minimum Wages provides information on minimum wage rates; all of these data are relevant for constructing a measure of minimum wage bindingness.

The Centers for Disease Control (CDC) National Vital Statistics System offers data on total drug poisoning mortality by county, while the CDC's Multiple Cause of Death database provides overdose deaths at the state-year level for different drugs.
Finally, the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) offers opioid distribution data.
The joint panel covers the United states for the period 2003-2016.

@@

Therefore, this paper sheds light on how labor market regulations, and in particular minimum wage rates, give rise to heterogeneities in the effects of the opioid crisis on the labor market, contributing to a better understanding of the ongoing crisis and the policies implemented to address it.
More broadly, it contributes to the understanding of how labor market regulations interact with drug disorders.

The paper is structured as follows.
First, it provides a summary of the opioid crisis and the effects it has had on the labor market as uncovered by the literature.
Then, it discusses some of the policies implemented to curb the opioid misuse and its effectiveness.
Later, I describe how different minimum wage rates can explain some of these policy consequences and the heterogeneity of the labor market outcomes. 
A model is introduced to formalize the intuition behind this argument.
The employed data is discussed before presenting the empirical strategy and obtained results.

@@

% CHECKED
In this section, I summarize the origins of the opioid crisis and its general effects on the labor market.
I then discuss various policies implemented to curb opioid misuse, and how some of them resulted ineffective or counterproductive. 
Finally, I explain how labor market incentives, in particular minimum wage rates, can account for some of these policy failures and the heterogeneity in labor market outcomes.

@@

% CHECKED
Although the impact of minimum wages on labor demand has been questioned since \textcite{Card1994} published their seminal paper, recent research indicates that minimum wage increases can negatively affect employment in certain contexts.
For instance, we know that minimum wage increases can lead to a decrease in employment in tradeable sectors (\cite{Cengiz2019}), and in industries where passing costs to consumers is more difficult (\cite{Harasztosi2019}).
This impact seems to be more pronounced for low-skilled workers (\cite{Neumark2004}), although the effect is negligible for small increases (\cite{Clemens2021}).
The macroeconomic situation can also influence the effect of minimum wage increases on employment if the former is particularly binding (\cite{Clemens2019}). 
However, \textcite{Engbom2022} noted that these effects can be muted by reallocations of workers across firms.
Crucially, they argue that lower-productivity firms reduce vacancy creation and hiring, which in our context could lead to lower job finding rates where the minimum wage is more binding.

@@

From (1), we see that the determinants of the decision to quit are crucial to understanding the effects of a prescription law.
This decision depends on the labor market opportunities available to the individual.
The addicted individual faces a labor market with a job finding rate

@@

The decision to quit opioids is hence influenced by the minimum wage as follows:
If the individual quits, she will seek a job that provides utility \(u(w_{it}L_{it})\), where \(w_{it}\) is the wage she receives and \(L_{it}\) is the number of hours she works.
The probability of finding a job is given by the job finding rate.
If the product of the utility and the job finding rate is greater than the disutility of quitting, \(\mathcal{C}^{quit}\), the individual will quit.
Formally

@@

Unless a corner solution exists (i.e., the disutility of quitting is greater or lower than the maximum utility the individual can get from working), there must be some wage level such that

@@

that is, for some wage level, the job finding rate ensures that the utility from working exceeds the disutility of quitting.
Notice that this is irrespective of the differences in levels of wages, since the job finding rate is also playing a role.

If condition (6) holds, after a prescription law is enacted (\(z_{t} = 1\)) we should observe an increase in participation rates where the minimum wage is less binding (\(O_{it} = 0\)) but also an increase in illegal opioid consumption where the minimum wage is more binding (\(O_{it} = h_{it}\)).
We test these hypotheses in the next sections.

@@

% CHECKED
To uncover the heterogeneity in the effects of the opioid crisis on the labor market created by minimum wage rates, I use several data sources.
The result is a panel of county-level data for the United States for the period 2003-2016 encompassing labor market, demographic and health variables.
Monthly data is utilized wherever available; otherwise, I use the data available for the year.
Similarly, state-level minimum wage rates are allocated to the counties.

% CHECKED
\textit{Labor market data and demographics}: 
I use the US Bureau of Labor Statistics (BLS) Local Area Unemployment Statistics (LAUS) database to access county-level labor market statistics. 
Specifically, I gather monthly data on labor force participation, employment, and unemployment rates at the month-county level.
This dataset covers the period 1990-2023. 
Additionally, through the US Census Bureau's Population Estimates Program (PEP), I obtain demographic information at the county-year level spanning from 1969 to 2020. 
This dataset provides the population counts of each county by gender, race and origin (Hispainc or Non-Hispanic), on a yearly basis.
Moreover, I can obtain the number of people in each of the 19 age groups considered, which is used to construct the labor force participation rate.

% CHECKED
\textit{Minimum wage bindingness}: 
I utilize the US Department of Labor's Changes in Basic Minimum Wages dataset to gather information on minimum wage rates at the state level from 1968 to 2023.
Estimates of the wage distribution at the county level are requiered to construct a measure of minimum wage bindingness, as outlined in the next section.
For this purpose, I rely on the US Bureau of Labor Statistics (BLS) Occupational Employment and Wage Statistics (OEWS) database to obtain national industry-specific occupations and wages statistics.
This dataset reports the number of workers in each occupation and the 10th, 25th, median, 75th and 90th wage percentiles for each industry, from 1997 to 2023.
Additionally, I use the data from \textcite{Eckert2020} to obtain county-level industry-specific employment data spanning from 1975 to 2016.
These last two datasets have an annual frequency. 4-digit NAICS codes are used to match the industries across datasets.

% CHECKED
\textit{Prescription Drug Monitoring Programs (PDMPs)}: 
I use data from \textcite{Horwitz2021} to obtain information on Prescription Drug Monitoring Programs (PDMPs) from 1990 to 2019.
This dataset offers state-year level detailing the implementation of PDMP types across states, as illustrated in Table 1. 
PDMPs are categorized into Modern System and Must Query types. 
A Modern System PDMP grants electronic database access to authorized users, whereas a Must Query PDMP mandates prescribers to consult the database before prescribing listed opioids. 
This last type of PDMP is the one most likely to have an impact on opioid misuse, and it forms the basis of my estimations.

% CHECKED
\textit{Prescriptions and overdose deaths}: 
I use data from the Centers for Disease Control (CDC) National Center for Health Statistics (NCHS) to obtain information on total drug poisoning mortality by county, from 2003 to 2021.
To overcome the suppression of data following anonimity concerns, \textcite{CLMort} employs Hierarchical Bayesian models with spatial and temporal random effects to estimate the number of deaths where the data is not available.
I further use the Multiple Cause of Death database (\cite{CDCMCD9920}), which reports overdose deaths at the state-year level for different drugs, from 1999 to 2020, to obtain information on overdose deaths related to heroin (T.40.1), methadone (T.40.2), other opioids (T.40.3) and other synthetic narcotics (T.40.4).
The codes reffer to the Multiple Causes of Death ICD-10 classification. 
Overdose deaths are identified using the Underlying Cause of Death ICD-10 codes X40-X44, X60-X64, X85, and Y10-Y14.

% CHECKED
Finally, I use the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) to obtain opioid distribution data spanning from 2006 to 2014.
This dataset records the distribution of all Schedule II substances by active ingredient at the 3-digit ZIP code level.
The data is aggregated on semi-synthetic opioid analgesics (i.e. oxycodone, hydrocodone) prescribed at the county-month level.
The shorter time span does not affect the main specifications, since I use prescription rates only to assure that the PDMPs are effective in a complementary exercise.

@@

Using the labor force participation rate as the dependent variable, the results of the main specification are displayed in Figure 2.
The figure shows the distribution of the coefficients of the impact of the passing of the Must Query PDMPs on labor force participation rates across different levels of minimum wage bindingness, for different periods relative to treatment.
In the upper right panel we can see the immediate impact of the policy.
For those counties where the minimum wage is less binding, the effect is positive (1.37 p.p.) and significant (0.513, 2.22) at conventional levels.
The effects become smaller and more uncertain as the minimum wage becomes more binding.
For those counties in which the minimum wage has a higher impact on labor demand, the point estimate becomes negative (-0.693) but not significant (-2.54, 1.16).

The effect seems to be persistent in the short run (upper right and lower right panels), but fades away after two years (lower right)\footnote{The confidence interval of the most binnding counties is not displayed in this last panel for visualization purposes. The values are (-3.10, 1.90)}.
Figure 3 shows the time averages of the coefficients for two years after the law is passed.
The results remain positive and significant for those counties in which the minimum wage is less binding (1.06 p.p., (0.584, 1.55)), while they are not significant for those where the minimum wage was more binding.

The results seem to be driven mostly by the impact of the policy on prescription rates.
Although the overlap between the passing of the policies (Table 1) and the data for prescription rates (2006-2014) is scarce, we can still obtain som estimates on the impact of Must Query PDMPs on prescription rates per capita.
Figure 4 runs the model (10) with the prescription rates per capita as the dependent variable.
The two months after the treatment is passed are displayed.
We see that the policy is able to decrease the amount of prescriptions, with the biggest drops occuring where the minimum wage is the least and the most binding (the effects in these areas are not statistically different). 
Although the target of these policies where not to decrease prescriptions \textit{per se}, but only tackling misuse cases, the results suggest that the policies are effective in reducing the amount of opioids prescribed on the aggregate.

Moreover, this seems to be the main channel through which the policies affect labor market outcomes: Figure 8 in the appendix shows that the relation between labor market outcomes and minimum wage bindingness is not significant when controlling for prescription rates.
Figures 9 and 10 in the Appendix show the the point estimates of the binscatter for different Kaitz indices, for six months and two years after the policy is passed.
It is easy to see that the results do not change much across different measures, this being especially true the closer in time the treatment was implemented.

\begin{comment}
After six months:
1.79 (0.955, 2.63)
0.333 (-0.665, 1.33)

After two years:
-0.118 (-1.66, 1.42)
-0.599 (-3.10, 1.90)

Time averages:
1.06  (0.584, 1.55)
0.182 (-0.335, 0.698)
\end{comment}

@@

The analysis is repeated using the unemployment rate as the dependent variable.
The results are displayed in the same fashion in Figure 5.
The distribution of the coefficients reveal that almost all the positive effect it was found in the labor force participation rates translated into an increase in the unemployment rate.
For those counties where the minimum wage is less binding, the effect is positive (1.02 p.p.) and significant (1.46, 0.575) at conventional levels. 
The point estimate is only 0.35 p.p. below the one it was found for the labor force participation rate.
Although a positive effect is also found for those counties where the minimum wage is more binding (0.640 p.p.), the effect is not significant (-0.0252, 1.31).

These results shouldn't be surprising, at least in the short run: if people with severe opioid addictions start looking for a job after the implementation of a PDMP, they might not be able to find one immediately even in those counties where the minimum wage is less binding.
Strikingly though, the impact on unemployment persists after two years.
The lower right panel shows that, even after the increase in labor force participation rates fades away, the increase in unemployment rates remains high (2.77 p.p.) and significant (1.87, 3.68) where the minimum wage is less binding.
Whether the causes of this persistence are related to opioid consumption (through, say, relapsing) or to characterisctics of the workers or the labor market (say, if these workers face more frictional unemployment) remains an open question.

\begin{comment}
    After six months:
    1.01   1.48   0.543
    -0.0981 0.199 -0.396

    After two years:
    2.77   3.68   1.87
    -0.0720 0.265 -0.409 
\end{comment}

The time average of the coefficients, displayed in Figure 6, confirm our previous result.
The effect of the policy on unemployment rates is positive and significant for those counties where the minimum wage is less binding (1.25 p.p., (0.97, 1.52)), while it is not significant for those where the minimum wage is more binding (0.025 p.p., (-0.11, 0.16)).
Again, the results seem to be driven by the impact of the policy on prescription rates, especially in the medium run: Figure 11 in the Appendix shows that the impact stops being significantly different from zero after one year once we control for prescriptions per capita.
Figures 12 and 13 in the Appendix show the binscatter for different Kaitz indices, for six months and two years after the policy is passed.
The results do not change substantially across different measures.

@@

Finally, and in accordance with the model outlined above, I show some evidence that the rollout of Must Query PDMPs has had a positive impact on the consumption of illegal opioids in those counties where the minimum wage was more binding.
For this purpose, I run the same estimation strategy as before, but using the number of overdose deaths per capita for different opioids as the dependent variable. 
The data is at the state-year level, and the number of total poisoning deaths is not used as a control variable.
Figure 7 displays the scatter plot of the time average of the coefficients on the Kaitz-0.10 index of the states, for the two years after the policy is passed.
The opioids included are heroin, methadone, other opioids and other synthetic narcotics, following the ICD-10 classification.

The coefficients are regressed on the Kaitz index and a constant. 
Tables 2-5 show the values of the regression coefficients for each of the different opioids.
Although admittedly weaker than the evidence presented in previous sections due to data constaints, it is interesting that the only marginally significant relation is found for heroin.
This suggest that the positive relation between black market switching and low labor demand is present, but we cannot assert this relationship with the same degree of certainty.

It seems implausible that part of the effect is absorbed by the consumption of other opioids.
Methadone is often used as a treatment for opioid adiction and can be legally used under a doctor's supervision.
Other opioids include drugs such as morphine and codeine, but also semi-synthetic opioid analgesics (e.g. drugs such as oxycodone, hydrocodone, hydromorphone, and oxymorphone), some of them targeted by the PDMPs.
Finally, although the category of other synthetic narcotics includes fentanyl and its analogues, its use didn't used to be as widespread as it is today during the period considered.

\section*{Conclusion}

This paper contributes to the understanding of how labor market regulations, particularly minimum wage rates, shape the relationship between opioid misuse and labor market outcomes. By leveraging the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, this study provides a nuanced view of how labor market opportunities influence the incentives for opioid users to quit or switch to illegal alternatives.

The findings reveal that labor force participation and unemployment rates increased as a result of prescription policies only in areas where the minimum wage was not binding. In contrast, a surge in heroin consumption occurred where the minimum wage was binding, highlighting the complex interplay between labor market conditions and opioid misuse. These results underscore the importance of considering labor market regulations when evaluating the economic impacts of the opioid crisis.

Furthermore, this study adds to the growing body of literature that examines the economic effects of the opioid epidemic. While previous research has primarily focused on the direct consequences of opioid misuse on labor market outcomes, this paper emphasizes the role of labor market regulations in mediating these effects. The heterogeneity in the impact of prescription policies across different levels of minimum wage bindingness suggests that policies aimed at curbing opioid misuse must account for the broader economic context to be effective.

In summary, the analysis demonstrates that higher minimum wages, by reducing labor demand for low-skilled workers, can diminish the incentives for opioid users to quit, thereby exacerbating the shift towards illegal alternatives like heroin. These insights highlight the need for a comprehensive approach to addressing the opioid crisis, one that integrates labor market policies with public health interventions to effectively mitigate the epidemic's adverse effects on both health and economic outcomes.

Future research should continue to explore the interplay between labor market regulations and health crises, considering the potential unintended consequences of well-intentioned policies. By doing so, policymakers can design more effective interventions that address the root causes of public health crises while also supporting economic stability and growth.

[MOVING FORWARD]

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{lcccc}
    \toprule
     & \textbf{Estimate} & \textbf{Std. Error} & \textbf{t value} & \textbf{Pr($>|t|$)} \\
    \midrule
    Intercept  & 5.600 & 2.135 & 2.623 & 0.0211 * \\
    Kaitz-0.10 & 9.613 & 5.021 & 1.914 & 0.0778 . \\
    \midrule
    \multicolumn{5}{l}{\textit{Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1}} \\
    \midrule
    \multicolumn{5}{l}{\textbf{Residual standard error}: 1.95 on 13 degrees of freedom} \\
    \multicolumn{5}{l}{\textbf{Multiple R-squared}: 0.2199, \textbf{Adjusted R-squared}: 0.1599} \\
    \multicolumn{5}{l}{\textbf{F-statistic}: 3.665 on 1 and 13 DF, \textbf{p-value}: 0.07783} \\
    \bottomrule
    \end{tabular}
    \caption{Regression Results: Heroin Overdose Deaths}
    \label{tab:regression_results_heroin}
\end{table}

\begin{table}[ht]
    \centering
    \begin{tabular}{lcccc}
    \toprule
     & \textbf{Estimate} & \textbf{Std. Error} & \textbf{t value} & \textbf{Pr($>|t|$)} \\
    \midrule
    Intercept  & -0.4800 & 0.8021 & -0.598 & 0.560 \\
    Kaitz-0.10 & -0.6460 & 1.8900 & -0.342 & 0.738 \\
    \midrule
    \multicolumn{5}{l}{\textit{Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1}} \\
    \midrule
    \multicolumn{5}{l}{\textbf{Residual standard error}: 0.7576 on 13 degrees of freedom} \\
    \multicolumn{5}{l}{\textbf{Multiple R-squared}: 0.008906, \textbf{Adjusted R-squared}: -0.06733} \\
    \multicolumn{5}{l}{\textbf{F-statistic}: 0.1168 on 1 and 13 DF, \textbf{p-value}: 0.738} \\
    \bottomrule
    \end{tabular}
    \caption{Regression Results: Methadone Overdose Deaths}
    \label{tab:regression_results_methadone}
\end{table}

\begin{table}[ht]
    \centering
    \begin{tabular}{lcccc}
    \toprule
     & \textbf{Estimate} & \textbf{Std. Error} & \textbf{t value} & \textbf{Pr($>|t|$)} \\
    \midrule
    Intercept  & -3.486 & 2.530 & -1.378 & 0.192 \\
    Kaitz-0.10 & -7.537 & 5.949 & -1.267 & 0.227 \\
    \midrule
    \multicolumn{5}{l}{\textit{Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1}} \\
    \midrule
    \multicolumn{5}{l}{\textbf{Residual standard error}: 2.31 on 13 degrees of freedom} \\
    \multicolumn{5}{l}{\textbf{Multiple R-squared}: 0.1099, \textbf{Adjusted R-squared}: 0.04143} \\
    \multicolumn{5}{l}{\textbf{F-statistic}: 1.605 on 1 and 13 DF, \textbf{p-value}: 0.2274} \\
    \bottomrule
    \end{tabular}
    \caption{Regression Results: Other opioids Overdose Deaths}
    \label{tab:regression_results_otheropioids}
\end{table}

\begin{table}[ht]
    \centering
    \begin{tabular}{lcccc}
    \toprule
     & \textbf{Estimate} & \textbf{Std. Error} & \textbf{t value} & \textbf{Pr($>|t|$)} \\
    \midrule
    Intercept  & 4.944 & 5.995 & 0.825 & 0.424 \\
    Kaitz-0.10 & 8.056 & 14.097 & 0.571 & 0.577 \\
    \midrule
    \multicolumn{5}{l}{\textit{Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1}} \\
    \midrule
    \multicolumn{5}{l}{\textbf{Residual standard error}: 5.475 on 13 degrees of freedom} \\
    \multicolumn{5}{l}{\textbf{Multiple R-squared}: 0.0245, \textbf{Adjusted R-squared}: -0.05053} \\
    \multicolumn{5}{l}{\textbf{F-statistic}: 0.3266 on 1 and 13 DF, \textbf{p-value}: 0.5774} \\
    \bottomrule
    \end{tabular}
    \caption{Regression Results: synthetic opioids Overdose Deaths}
    \label{tab:regression_results_synthetic}
\end{table}
